An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-10-12 aistats2017.tex
OK it is now 8 pages. Still need Xinlei's Bardenet results.

Our test in Equation~(\ref{eq:criteria}) requires knowing the distribution of
$X_{\rm corr}$. In Section~\ref{sec:analysis}, we show that the test accuracy
depends on the absolute error between the CDFs of $X_{\rm norm} + X_{\rm corr}$
and $X_{\rm log}$. Consequently, we need to minimize this in our construction of
$X_{\rm corr}$. More formally, let $\Phi_{s_X} = \Phi(X/s_X)$ where $\Phi$ is
the standard normal CDF\footnote{Hence, $\Phi_{s_X}$ is the CDF of a zero-mean
Gaussian with standard deviation $s_X$.}, $S(X)$ be the logistic function, and
$C_{\sigma}(X)$ be the \emph{density} of the correction $X_{\rm corr}$
distribution. Our goal is to solve:
\begin{equation}\label{eq:overall_corr_problem}
    C_\sigma^* = \argmin_{C_\sigma} |\Phi_{\sigma} * C_{\sigma} - S|
\end{equation}
where $*$ denotes convolution.

For computation of $C_\sigma$, we assume that its input $Y$ and another variable
$X$ lie in the intervals $[-V,V]$ and $[-2V,2V]$, respectively.
% Daniel: removing this because we were explaining this in quite some detail and
% it was taking a lot of space.
%\[
%    \argmin_{C\sigma} \sup_{x \in [-2V,2V]}\left|\int_{-V}^{V}\Phi_{\sigma}(x-y) C_{\sigma}(y)dy - S(x)\right|.
%\]
We discretize the convolution by discretizing $X$ and $Y$ into $4N+1$ and $2N+1$
values respectively. If $i \in \{-2N, \ldots, 2N\}=\mathcal{I}$ and $j \in \{-N,
\ldots, N\}=\mathcal{J}$, then we can write $X_i = i(V/N)$ and $Y_j = j(V/N)$, and the
objective can be written as:
\[
C_\sigma^* = \argmin_{C_\sigma} \max_{i \in \mathcal{I}}\left|\sum_{j\in\mathcal{J}} \Phi_{\sigma}(X_i-Y_j) C_{\sigma}(Y_j) - S(X_i)\right|.
\]
We now define a matrix $M$ and vectors $u$ and $v$ such that $M_{ij} =
\Phi_{\sigma}(X_i-Y_j)$, $u_j = C_{\sigma}(Y_j)$, and $v_i = S(X_i)$, where the
indices $i$ and $j$ are appropriately translated to be non-negative indices for
$M, u,$ and $v$. Thus, the problem is now to minimize $\|Mu-v\|_{\infty}$ with
the constraint that $u > 0$ since it represents a density. We approximate this
with a least squares solution:
\begin{equation}\label{eq:optimization_l2}
    u^* = \argmin_u\; \|Mu-v\|_2^2 + \lambda \|u\|_2^2,
\end{equation}
with regularization $\lambda$. The solution is well-known from the normal
equations ($u^* = (M^TM + \lambda I)^{-1}M^Tv$) and in practice yields an
acceptable $L_{\infty}$ norm.

@@

\begin{itemize}[noitemsep]
    \item It uses an adaptive step size so as to use the smallest possible
    average minibatch size. Unlike previous work however (and as we show in
    Section~\ref{sec:experiments}) the size distribution is short-tailed.

@@

    \item The sample variance $s^2_{\Delta^*}$ is proportional to
    $\|\theta'-\theta\|_2^2$ whose distribution for Normal proposals is the
    square of a normal variable. 
\end{itemize}

@@

We now derive error bounds for our M-H test and the approximate target
distribution it generates. In the most similar prior
works,~\citet{cutting_mh_2014} uses asymptotic arguments based on the CLT to
argue that its approximate acceptance test error tends to zero as batch size
increases, but no quantitative bounds are given.
In~\citet{icml2014c1_bardenet14}, explicit bounds are given, but they depend on
bounding:
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i\mid\theta') - \log p(x_i\mid\theta)|.
\end{equation}
It is possible to analytically determine $C_{\theta,\theta'}$ for logistic
regression, but it is unclear how to do so for more complicated models, and one
generally needs to use all $p(x_i\mid \theta')$ terms\footnote{The sample code
provided by the authors of~\citet{icml2014c1_bardenet14} in fact computes
$C_{\theta,\theta'}$ explicitly for each sample generated, i.e. it traverses the
entire data, thus providing no performance advantage over the complete test.} to
get $C_{\theta,\theta'}$.  In contrast, we use quantitative forms of the Central
Limit Theorem which rely on measurable statistics from a single minibatch. Thus
a sampler using our approach does not need to see data beyond the current
minibatch.

@@

\begin{lemma}\label{lem:quant_clt}
Let $X_1,\ldots,X_n$ be a set of zero-mean, independent, identically-distributed
random variables with sample mean $\bar{X}$ and sample variance $s^2_X$
where:
\begin{equation}
    \bar{X} = \frac{1}{n}\sum_{i=1}^nX_i, \quad s_X = \frac{1}{n}\left(\sum_{i=1}^n(X_i-\bar{X})^2\right)^{\frac{1}{2}}.
\end{equation}
This implies $t=\bar{X}/s_X$ has an approximate Student's distribution
which approaches a normal distribution in the limit. Then
\begin{equation}\label{eq:clt-bounds}
    \sup_x|{\rm Pr}(t<x) - \Phi(x)| \leq \frac{6.4E|X|^3+2E|X|}{\sqrt{n}}.
\end{equation}
\end{lemma}

@@

Lemma~\ref{lem:quant_clt} demonstrates that as long as we know the first and
third absolute moments $E|X|$ and $E|X|^3$, we can bound the error of the normal
approximation, which decays as $O(n^{-\frac{1}{2}})$. Making the change of
variables $y = x s_X$, Equation~(\ref{eq:clt-bounds}) becomes
\begin{equation}\label{eq:clt-bounds2}
   \sup_y\left|{\rm Pr}(\bar{X}< y) - \Phi\left(\frac{y}{s_X}\right)\right| \leq \frac{6.4E|X|^3+2E|X|}{\sqrt{n}}
\end{equation}
showing that the distribution of $\bar{X}$ approaches the normal distribution
$\mathcal{N}(0,s_X)$ whose variance is $s_X$,
measured from the sample.

@@

From Section~\ref{sec:our_algorithm}, as the distribution of $\Delta^*$
approaches a Gaussian, our new MH test becomes more accurate.
Corollary~\ref{cor:bounds_preserved} shows that the bounds from
Section~\ref{ssec:delta_star_distribution} are preserved after the addition of
the random variables we use, showing that our test should remain accurate.
% Daniel: I'm moving this paragraph over here and putting the rest of it in the
% supplementary material since our code doesn't do this, so we should not
% describe this in much detail.
In fact we can do better ($O(n^{-1})$) by using a more precise limit
distribution under an additional assumption. We review this in the Supplementary
Material, Section~\ref{app:better_error_bound}, and leave the details to future
work.

@@

We conduct two sets of experiments analyzing convergence to the posterior of a
heated Gaussian mixture, and efficiency on logistic regression.

% Daniel: moving this here so figures don't overflow to next pages. We can tune
% the positioning if needed.
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.85\linewidth]{GaussianMixtureResult/posterior_of_gaussian.png}
    \caption{
    The log posterior contours and scatter plots of sampled $\theta$ values
    using different methods. 
    }
    \label{fig:gauss_mix_1}
\end{figure*}
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.85\linewidth]{minibatch_size_gaussian.png}
    \caption{
    Minibatch sizes used in Section~\ref{ssec:gaussians}'s experiment. The axes
    have the same (log-log scale) range.
    }
    \label{fig:gauss_mix_2}
\end{figure*}

@@

We borrow a Gaussian mixture experiment from~\citet{langevin_2011}.  The
parameter is 2-D, $\theta = \langle \theta_1,\theta_2 \rangle$, and the
generation process is
\begin{equation}\label{eq:data_generation}
\begin{split}
    \theta &\sim \mathcal{N}(0, {\rm diag}(\sigma_1^2,\sigma_2^2)) \\
    x_i & \sim 0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{split}
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$.  We fix $\theta =
\langle 0,1 \rangle$. The original paper sampled 100 data points and estimated
the posterior. We are interested in performance on larger problems and so
sampled 1,000,000 points to form the posterior of
$p(\theta)\prod_{i=1}^{1,000,000}p(x_i\mid \theta)$ with the same prior from
Equation~(\ref{eq:data_generation}). This produces a much sharper posterior with
two very narrow peaks.  Our goal is to reproduce the original posterior, so we
adjust the temperature to $T=10,000$.  Taking logs, we get the target as shown
in the far left of Figure~\ref{fig:gauss_mix_1}.

@@

Figure~\ref{fig:gauss_mix_2}, however, suggests that our method dominates in
terms of speed and efficiency. The histograms of the (final) minibatch sizes
used each iteration show that our method consumes significantly less data; the
distribution is short-tailed and the mean is 210. The other methods occasionally
need to use nearly all data points, and average minibatch sizes are 15562 and
16857. The sizes roughly predict the running times of these methods, which are
proportional to the data consumed, with the exception
of~\citet{icml2014c1_bardenet14}, which requires a full pass to compute
$C_{\theta,\theta'}$ (Equation~(\ref{eq:bad_bound})). Using minibatch sizes
as a proxy for time also avoids discrepancies due to code optimization.

% Daniel: moving this here so figures don't overflow to next pages. We can tune
% the positioning if needed.
\begin{figure*}[t]
	\centering
	%\includegraphics[width=0.9\linewidth]{logistic_performance.png}
	\includegraphics[width=0.85\linewidth]{logistic_regression_ll_acc_results.png}
	\caption{
    Logistic regression performance (accuracy/log likelihood) based on
    cumulative data usage.
    }
	\label{fig:logistic_performance}
\end{figure*}
\begin{figure*}[t]
	\centering
	%\includegraphics[width=0.9\linewidth]{minibatch_size_logistic.png}
	\includegraphics[width=0.85\linewidth]{logistic_regression_histograms_results.png}
	\caption{
    Counts of minibatch sizes in the MNIST logistic regression experiment
    (analogous to Figure~\ref{fig:gauss_mix_2}).
    }
	\label{fig:logistic_minibatch}
\end{figure*}

@@

Figure~\ref{fig:logistic_performance} shows the test log likelihood and
prediction accuracy as a function of the cumulative training points
processed.\footnote{Note that the curves do not span the same length over the
x-axis, because our test consumes fewer samples throughout the MCMC procedure,
so the corresponding curve will ``end'' before the other two.} To generate the
curves, for each of the sampled vectors $\theta_t$, $t\in\{1,\ldots,3000\}$, we
use $\theta_t$ as the parameter for logistic regression.  Our minibatch MH test
is more efficient in terms of accuracy, achieving convergence using roughly half
as many elements compared to~\citet{cutting_mh_2014} and \textbf{TODO TODO TODO}
compared to~\citet{icml2014c1_bardenet14}, though its log likelihood results lag
slightly behind the former algorithm during the very early stages.

@@

In this paper, we derived an M-H test for minibatch MCMC which approximates full
data tests. We provide theoretical results and experimentally show the benefits
of our test on Gaussian mixtures and logistic regression. Directions for future
work include running more experiments with a particular focus on controlling
variances.  More elaborate extensions include testing on neural networks,
combining our results with Hamiltonian Monte Carlo methods, providing a recipe
for how to use our algorithm (following the framework of~\citet{sgmcmc_2015}),
or integrating parallel MCMC~\citep{conf/uai/AngelinoKWSA14,conf/icml/AhnSW14}
concepts.

>>>>

Our test in Equation~(\ref{eq:criteria}) requires knowing the distribution of
$X_{\rm corr}$. In Section~\ref{sec:analysis}, we show that the test accuracy
depends on the absolute error between the CDFs of $X_{\rm norm} + X_{\rm corr}$
and $X_{\rm log}$. Consequently, we need to minimize this in our construction of
$X_{\rm corr}$. More formally, let $\Phi_{s_X} = \Phi(X/s_X)$ where $\Phi$ is
the standard normal CDF\footnote{Hence, $\Phi_{s_X}$ is the CDF of a zero-mean
Gaussian with standard deviation $s_X$.}, $S(X)$ be the logistic function, and
$C_{\sigma}(X)$ be the \emph{density} of the correction $X_{\rm corr}$
distribution. Our goal is to solve:
\begin{equation}\label{eq:overall_corr_problem}
    C_\sigma^* = \argmin_{C_\sigma} |\Phi_{\sigma} * C_{\sigma} - S|
\end{equation}
where $*$ denotes convolution. To compute $C_\sigma$, we assume the input $Y$
and another variable $X$ lie in the intervals $[-V,V]$ and $[-2V,2V]$,
respectively.  We discretize the convolution by discretizing $X$ and $Y$ into
$4N+1$ and $2N+1$ values respectively. If $i \in \{-2N, \ldots,
2N\}=\mathcal{I}$ and $j \in \{-N, \ldots, N\}=\mathcal{J}$, then we can write
$X_i = i(V/N)$ and $Y_j = j(V/N)$, and the objective can be written as:
\[
C_\sigma^* = \argmin_{C_\sigma} \max_{i \in \mathcal{I}}\left|\sum_{j\in\mathcal{J}} \Phi_{\sigma}(X_i-Y_j) C_{\sigma}(Y_j) - S(X_i)\right|.
\]
Now define matrix $M$ and vectors $u$ and $v$ such that $M_{ij} =
\Phi_{\sigma}(X_i-Y_j)$, $u_j = C_{\sigma}(Y_j)$, and $v_i = S(X_i)$, where the
indices $i$ and $j$ are appropriately translated to be non-negative for $M, u,$
and $v$. The problem is now to minimize $\|Mu-v\|_{\infty}$ with the density
non-negative constraint $u > 0$. We approximate this with least squares:
\begin{equation}\label{eq:optimization_l2}
    u^* = \argmin_u\; \|Mu-v\|_2^2 + \lambda \|u\|_2^2,
\end{equation}
with regularization $\lambda$. The solution is well-known from the normal
equations ($u^* = (M^TM + \lambda I)^{-1}M^Tv$) and in practice yields an
acceptable $L_{\infty}$ norm.

@@

\begin{itemize}[noitemsep]
    \item It uses an adaptive step size so as to use the smallest possible
    average minibatch size. Unlike previous work, the size distribution is
    short-tailed.

@@

    % Daniel: I cut some space to use this:
    \item The sample variance is denoted as $s^2_{\Delta^*}$ and is proportional
    to $\|\theta'-\theta\|_2^2$.

    % Daniel: instead of this:
    %\item The sample variance $s^2_{\Delta^*}$ is proportional to
    %$\|\theta'-\theta\|_2^2$ whose distribution for Normal proposals is the
    %square of a normal variable. 
\end{itemize}

@@

We now derive error bounds for our M-H test and the target distribution it
generates. In the most similar prior works,~\citet{cutting_mh_2014} use CLT
asymptotic arguments to show that its approximate acceptance test error tends to
zero as batch size increases, but no quantitative bounds are given.
In~\citet{icml2014c1_bardenet14}, explicit bounds are given and depend on
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i\mid\theta') - \log p(x_i\mid\theta)|.
\end{equation}
It is possible to analytically determine $C_{\theta,\theta'}$ for logistic
regression, but it is unclear how to do so for more complicated models, and one
generally needs to use all $p(x_i\mid \theta')$ terms\footnote{The sample code
provided by~\citet{icml2014c1_bardenet14} computes $C_{\theta,\theta'}$
by traversing the entire data, thus providing no performance advantage over the
complete test.} to get $C_{\theta,\theta'}$.  In contrast, we use quantitative
forms of the CLT which rely on measurable statistics from a \emph{single}
minibatch.

@@

\begin{lemma}\label{lem:quant_clt}
Let $X_1,\ldots,X_n$ be a set of zero-mean, independent, identically-distributed
random variables with sample mean $\bar{X}$ and sample variance $s^2_X$
where:
\begin{equation}
    \bar{X} = \frac{1}{n}\sum_{i=1}^nX_i, \quad s_X = \frac{1}{n}\left(\sum_{i=1}^n(X_i-\bar{X})^2\right)^{\frac{1}{2}}.
\end{equation}
This means $t=\bar{X}/s_X$ has an approximate Student's distribution
which approaches a Gaussian. Then
\begin{equation}\label{eq:clt-bounds}
    \sup_x|{\rm Pr}(t<x) - \Phi(x)| \leq \frac{6.4E|X|^3+2E|X|}{\sqrt{n}}.
\end{equation}
\end{lemma}

@@

Lemma~\ref{lem:quant_clt} demonstrates that if we know $E|X|$ and $E|X|^3$, we
can bound the error of the normal approximation, which decays as
$O(n^{-\frac{1}{2}})$. Making the change of variables $y = x s_X$,
Equation~(\ref{eq:clt-bounds}) becomes
\begin{equation}\label{eq:clt-bounds2}
   \sup_y\left|{\rm Pr}(\bar{X}< y) - \Phi\left(\frac{y}{s_X}\right)\right| \leq \frac{6.4E|X|^3+2E|X|}{\sqrt{n}}
\end{equation}
showing that the distribution of $\bar{X}$ approaches the normal distribution
$\mathcal{N}(0,s_X)$ whose variance is $s_X$,
measured from the sample.

@@

Corollary~\ref{cor:bounds_preserved} shows that the bounds from
Section~\ref{ssec:delta_star_distribution} are preserved after adding 
random variables, so our test remains accurate.  In fact we can do better
($O(n^{-1})$ instead of $O(n^{-1/2})$) by using a more precise limit
distribution under an additional assumption. We review this in the Supplementary
Material, Section~\ref{app:better_error_bound}.

% Daniel: I moved the previous Section 5.4 part to the supplementary material
% since our code doesn't do this, so we should not describe this in much detail.

@@

% Daniel: moving this here so figures don't overflow to next pages. We can tune
% the positioning if needed.
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.8\linewidth]{GaussianMixtureResult/posterior_of_gaussian.png}
    \caption{
    The log posterior contours and scatter plots of sampled $\theta$ values
    using different methods. 
    }
    \label{fig:gauss_mix_1}
\end{figure*}
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.8\linewidth]{minibatch_size_gaussian.png}
    \caption{
    Minibatch sizes used in Section~\ref{ssec:gaussians}'s experiment. The axes
    have the same (log-log scale) range.
    }
    \label{fig:gauss_mix_2}
\end{figure*}

@@

% Daniel: moving this here so figures don't overflow to next pages. We can tune
% the positioning if needed.
\begin{figure*}[t]
	\centering
	%\includegraphics[width=0.9\linewidth]{logistic_performance.png}
	\includegraphics[width=0.8\linewidth]{logistic_regression_ll_acc_results.png}
	\caption{
    Logistic regression performance (accuracy/log likelihood) based on
    cumulative data usage.
    }
	\label{fig:logistic_performance}
\end{figure*}
\begin{figure*}[t]
	\centering
	%\includegraphics[width=0.9\linewidth]{minibatch_size_logistic.png}
	\includegraphics[width=0.8\linewidth]{logistic_regression_histograms_results.png}
	\caption{
    Counts of minibatch sizes in the MNIST logistic regression experiment
    (analogous to Figure~\ref{fig:gauss_mix_2}).
    }
	\label{fig:logistic_minibatch}
\end{figure*}

We first borrow a Gaussian mixture experiment from~\citet{langevin_2011}.  The
parameter is 2-D, $\theta = \langle \theta_1,\theta_2 \rangle$, and the
generation process is
\begin{equation}\label{eq:data_generation}
\begin{split}
    \theta &\sim \mathcal{N}(0, {\rm diag}(\sigma_1^2,\sigma_2^2)) \\
    x_i & \sim 0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{split}
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$.  We fix $\theta =
\langle 0,1 \rangle$. The original paper sampled 100 data points and estimated
the posterior. We are interested in performance on larger problems and so
sampled 1,000,000 points to form the posterior of
$p(\theta)\prod_{i=1}^{1,000,000}p(x_i\mid \theta)$ with the same prior from
Equation~(\ref{eq:data_generation}). This produces a much sharper posterior with
two very narrow peaks.  Our goal is to reproduce the original posterior, so we
adjust the temperature to $T=10,000$.  Taking logs, we get the target as shown
in the far left of Figure~\ref{fig:gauss_mix_1}.

@@

Figure~\ref{fig:gauss_mix_2}, however, suggests that our method dominates in
terms of speed and efficiency. The histograms of the (final) minibatch sizes
used each iteration show that our method consumes significantly less data; the
distribution is short-tailed and the mean is 210. The other methods occasionally
need to use nearly all data points, and average minibatch sizes are 15562 and
16857. The sizes roughly predict the running times of these methods, which are
proportional to the data consumed, with the exception of the
$C_{\theta,\theta'}$ time for~\citet{icml2014c1_bardenet14}.Using minibatch
sizes as a proxy for time also avoids discrepancies due to code optimization.

@@

Figure~\ref{fig:logistic_performance} shows the test log likelihood and
prediction accuracy as a function of the cumulative training points
processed.\footnote{The curves do not span the same length over the x-axis since
the methods consume different amounts of data.} To generate the curves, for each
of the sampled vectors $\theta_t$, $t\in\{1,\ldots,3000\}$, we use $\theta_t$ as
the parameter for logistic regression.  Our minibatch MH test is more efficient
in terms of accuracy, achieving convergence using roughly half as many elements
compared to~\citet{cutting_mh_2014} and \textbf{TODO TODO TODO} compared
to~\citet{icml2014c1_bardenet14}, though its log likelihood results lag slightly
behind the former algorithm during the very early stages.

@@

We have derived an M-H test for minibatch MCMC which approximates full data
tests. We present theoretical results and experimentally show the benefits of
our test on Gaussian mixtures and logistic regression. Directions for future
work include running more experiments with a particular focus on controlling
variances, testing on neural networks, combining our results with Hamiltonian
Monte Carlo methods, providing a recipe for how to use our algorithm (following
the framework of~\citet{sgmcmc_2015}), or integrating parallel
MCMC~\citep{conf/uai/AngelinoKWSA14,conf/icml/AhnSW14} concepts.
